Leadership across cultures and contexts: an exploratory analysis

### 2020-05-12 leadership_across_cultures_contexts.Rmd

```{r elasticcoercive, fig.width=10, fig.cap="Coefficients of logistic elasticnet regression models [@friedman_regularization_2010] of evidence for *Coercive authority*. A: Coefficients indicate the words whose frequencies in each text record best predicted evidence for *Coercive authority* in each text record. B: Non-zero coefficients of all quality and function variables that best predicted evidence for *Coercive authority*. Both models used the lasso penalty ($\\alpha=1$) and chose $\\lambda_{1se}$ by cross-validation (1 SE from $\\lambda_{min}$). Red values are positive predictors; purple values are negative predictors. Variables with coefficents = 0 not displayed. X-axes are odds ratios on a log 10 scale."}

@@

@@

@@

**Dominance and prestige**. Much of the evolutionary theoretical literature on leadership has suggested social influence is most often achieved via one of two distinct strategies: dominance or prestige [for review see @garfield_review_2019; @garfield_evolutionary_2019]. There is a long history of leadership styles being described as based on either force and strength (i.e., dominance) or respect and special skills (i.e., prestige). Such dual-models are rooted in the coupling of ethology and animal dominance hierarchies with ethnography and cultural prestige systems [see @mead_sex_1935; @sahlins_poor_1963; @lewis_leaders_1974; @tiger_dominance_1970; @bernstein_dominance_1976]. Synthesizing the ethnographic and ethological evidence, @kracke_force_1978 linked "force and persuassion" to psychological-affect between leaders and followers in small-scale societies and @barkow_prestige_1980 suggested the evolution of human prestige was linked to sexual selection and symbolic culture. The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and social hierarchy and inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019; @garfield_investigating_2019]. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on distinctions between dominance and prestige suggests, "human social stratification is characterized by the co-occurrence of both strategies." Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities, beyond dominance and prestige.

<!-- DISCUSSES COMPETENCIES FEATURE -->
**Leader capacities**. The individual-level competencies feature was comprised of variables associated with dominance-based leadership, including coercive authority and aggressiveness [@tiger_imperial_1971; @barkow_prestige_1975; @barkow_darwin_1989; @henrich_evolution_2001; @cheng_two_2013] as well as the qualities of knowledgeable or intelligent and experienced or accomplished, traits that have been emphasized by the prestige theory [@henrich_evolution_2001; @cheng_two_2013] and the more general computational services theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and which are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)). Interestingly, being feared and having supernatural capacities clustered more closely with the prestige-style traits than with the dominance-style traits. There is likely a strong connection between shamanism and ritual-based influence with both prestige and dominance styles of leadership. Ritual and supernatural capacities may represent a dimension of leader qualities in which dominance and prestige are coupled forming an alternative style of leadership, as shamans often provide valued public good, have special expertise and abilities, are feared, and achieve or maintain influence via threats or use of direct and/or indirect aggression. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate the superhuman abilities of shamans [@singh_cultural_2017].

As brief examples of the diversity of leader capacities, of the very many in the ethnographic record, among the Azande, "though the most knowledgeable witch-doctors tend to become leaders, there is no institutionalized organization or leadership among them" [@fo07-056, p. 85]; and among the Highland Scots, @es10-003 [p. 147] notes, "on the whole, their long ties with place and people, coupled with education, made them useful leaders of a community in which leaders were needed to adapt life to a changing economy." Such exceptionally talented leaders, however, also face unique challenges, particularly within more egalitarian societies which emphasizes personal autonomy and resit hierarchy. @nm09-056 [p. 183] describes Laura Cornelius Kellogg (also known as Minnie and Wynnogene, 1880-1947), a land-claims activist, writer, and perhaps the most famous female leader among the Iroquois:

<!-- DISCUSSES PROSOCIALITY FEATURE -->
**Prosocial norms**. The prosociality feature of leader qualities, including generosity, interpersonal and oratory skills, and charisma as well as the role of cultural norms in shaping leadership styles, including proper behavior and having a favorable personality, generally correspond to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @grabo_charisma_2017; @howell_role_2005]. The importance of prosociality  also corresponds to the @henrich_big_2015 Big-man mechanism, which suggests positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership.
<!-- A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity). -->

@@

<!-- One interesting pattern that has not been widely discussed in the evolutionary literature on leadership is the role of shamans and their intersection, in our results, with the dominant style of leadership. Our social costs cluster (Figure \@ref(fig:cluster)A) includes the *Supernatural* abilities variable. @winkelman1987cross found that in 47% of societies sampled from the HRAF ($n = 43$) shamans were associated with providing decision-making services and associated with military power. Providing ritual functions and military leadership are common (documented in over 50% of societies here) functions leaders serve (see Figure \@ref(fig:varsupport)). -->

>>>>

```{r elasticcoercive, fig.width=10, fig.cap="Coefficients of logistic elasticnet regression models of evidence for *Coercive authority*. A: Coefficients indicate the words whose frequencies in each text record best predicted evidence for *Coercive authority* in each text record. B: Non-zero coefficients of all quality and function variables that best predicted evidence for *Coercive authority*. Both models used the lasso penalty ($\\alpha=1$) and chose $\\lambda_{1se}$ by cross-validation (1 SE from $\\lambda_{min}$). Red values are positive predictors; purple values are negative predictors. Variables with coefficents = 0 not displayed. X-axes are odds ratios on a log 10 scale."}

@@

## Shaman leaders

```{r shamanstats}
supernatural_n <- sum(leader_text2$qualities_Supernatural==1)
shamanism_pct <- signif(100*sum(df_shaman$shamanism)/nrow(leader_text2), 3)
```

Shamans and other leaders with supernatural qualities were surprisingly common in the enthographic record. Of our `r nrow(leader_text2)` text records, `r sum(df_shaman$shaman)` used the word "shaman", `r supernatural_n` provided evidence for supernatural leader qualities, and `r sum(df_shaman$shamanism)` provided evidence for one or both (`r shamanism_pct`% of all text records). We term the latter variable *Shamanism*. We fit a logistic elasticnet model of shamanism as a function of all leadership dimensions (minus *Supernatural*). This model revealed that shaman leaders combined dimensions of dominant leaders (*Feared*) with dimensions of prestigious leaders (*Medical functions*, which are a form of knowledge or expertise, and *Experienced/accomplished*). See Figure \@ref(fig:shamanism).

```{r shamanism, fig.width=10, fig.cap="Coefficients of logistic elasticnet regression models of evidence for *Shamanism*. A: Coefficients indicate the words whose frequencies in each text record best predicted evidence for *Shamanism* in each text record. B: Non-zero coefficients of all quality and function variables that best predicted evidence for *Shamanism*. Both models used the lasso penalty ($\\alpha=1$) and chose $\\lambda_{1se}$ by cross-validation (1 SE from $\\lambda_{min}$). Red values are positive predictors; purple values are negative predictors. Variables with coefficents = 0 not displayed. X-axes are odds ratios on a log 10 scale."}
plot_shamanism_text <- plot_shamanism_text + scale_x_log10()

p <- plot_shamanism_text + plot_elastic_shamanism +
  plot_annotation(
    title = 'Predictors of Shamanism',
    tag_levels = 'A', 
    caption = 'Odds ratio', 
    theme=theme(plot.caption=element_text(hjust=0.5, size=12, margin=margin(t=0)))
    )
ggsave('Figures/shamanism_plot.pdf', p, width = 8, height = 4)
p
```

@@

@nm09-056 [p. 183] describes Laura Cornelius Kellogg (also known as Minnie and Wynnogene, 1880-1947), a land-claims activist, writer, and perhaps the most famous female leader among the Iroquois, whose life exemplified the risks to both leaders and followers:

@@

**Dominance and prestige**. Much of the evolutionary theoretical literature on leadership has suggested social influence is most often achieved via one of two distinct strategies: dominance or prestige [for review see @garfield_review_2019; @garfield_evolutionary_2019]. There is a long history of leadership styles being described as based on either force and strength (i.e., dominance) or respect and special skills (i.e., prestige). Such dual-models are rooted in the coupling of ethology and animal dominance hierarchies with ethnography and cultural prestige systems [see @mead_sex_1935; @sahlins_poor_1963; @lewis_leaders_1974; @tiger_dominance_1970; @bernstein_dominance_1976; @barkow_prestige_1975]. Synthesizing the ethnographic and ethological evidence, @kracke_force_1978 linked "force and persuassion" to psychological-affect between leaders and followers in small-scale societies and Barkow [-@barkow_prestige_1975; -@barkow_prestige_1980] suggested the evolution of human prestige was linked to intra- and inter-sexual selection of culturally acquired competencies.

@henrich_evolution_2001 introduced this important distinction to social science disciplines outside of anthropology, agreeing with @tiger_dominance_1970 and Barkow [-@barkow_prestige_1975; -@barkow_prestige_1980] on the evolutionary roots of dominance, but offering an evolutionary account of prestige rooted in social rather than sexual selection for culturally acquired competencies, which inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019; @garfield_investigating_2019]. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on distinctions between dominance and prestige suggests, "human social stratification is characterized by the co-occurrence of both strategies." Our analyses support the cross-cultural importance of leader aggression and coercion, as well as culturally acquired competencies, but also indicate multiple features of leader qualities beyond dominance and prestige.

<!-- DISCUSSES COMPETENCIES FEATURE -->
**Leader competencies and shamanism**. The individual competencies cluster (Figure \@ref(fig:cluster)) comprises variables associated with dominance-based leadership, including coercive authority and aggressiveness, as well as qualities associated with prestige-based leadership, including knowledgeable/intelligent and experienced/accomplished. These dimensions are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)).

Shamanism appears to be a distinct form of leadership that combines a strategy of inducing fear, similar to the dominance strategy, but is based on knowledge and expertise, similar to the prestige strategy. The *Competencies* feature included supernatural and feared dimensions (Figure \@ref(fig:cluster)). Our *Shamanism* variable (which was not one of our original leadership dimensions) was predicted by words that evoke dominance and prestige, such as *supernatural*, *power*, *fear*, and *knowledge* (Figure \@ref(fig:shamanism)A), and by four other leadership dimensions that also overlap with the dominance and prestige constructs: *Feared*, *Medicinal functions*, *Ritual functions*, and *Experience/accomplished*. Shamans often provide valued public good, have special expertise and abilities, are feared, and achieve or maintain influence via threats or use of direct and/or indirect aggression. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate the superhuman abilities of shamans [@singh_cultural_2017]. @ng06-054 describes shaman leaders among the Ojibwa:

>Leadership was vested in the heads of the co-residential groups who held their position through their hunting abilities and supernatural power as shamans. Shamans had the ability to foresee future events and the ability within certain limits to control them. These men were feared as well as respected and prior to 1900, most leaders were polygynists. Leadership in 1900, was based upon individual initiative and the vision quest and was not inherited, a fact that is congruent with the bilateral structure of Ojibwa society at this date (cf. Helm 1965).

<!-- As brief examples of the diversity of leader capacities, of the very many in the ethnographic record, among the Azande, "though the most knowledgeable witch-doctors tend to become leaders, there is no institutionalized organization or leadership among them" [@fo07-056, p. 85]; and among the Highland Scots, @es10-003 [p. 147] notes, "on the whole, their long ties with place and people, coupled with education, made them useful leaders of a community in which leaders were needed to adapt life to a changing economy." Such exceptionally talented leaders, however, also face unique challenges, particularly within more egalitarian societies which emphasizes personal autonomy and resit hierarchy. -->

<!-- DISCUSSES PROSOCIALITY FEATURE -->
**Prosocial norms**. The prosociality feature of leader qualities, including generosity, interpersonal and oratory skills, and charisma as well as the role of cultural norms in shaping leadership styles, including proper behavior and having a favorable personality (Figure \@ref(fig:cluster)), generally corresponds to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @grabo_charisma_2017; @howell_role_2005]. The importance of leader prosociality also supports the @henrich_big_2015 Big-man mechanism by which the positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership.
<!-- A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity). -->

@@

One interesting pattern that has not been widely discussed in the evolutionary literature on leadership is the role of shamans and their intersection, in our results, with the dominance and prestige styles of leadership. Our competencies cluster (Figure \@ref(fig:cluster)A) includes the *Supernatural* abilities dimension. @winkelman1987cross found that in 47% of societies sampled from the HRAF ($n = 43$) shamans were associated with providing decision-making services and associated with military power. Providing ritual functions and military leadership are common (documented in over 50% of societies here) functions leaders serve (see Figure \@ref(fig:varsupport)).
